Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks

### 2019-09-05 research/paper/paper.tex
Update paper draft

\begin{abstract}
Current speech emotion recognition systems often overlook the multi-label data that comes with databases. In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech. We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample. The model was able to classify the samples with an accuracy of 52.57\%. This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{\url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
\end{abstract}

@@

Current state-of-the-art speech emotion recognition has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN long short-term memory (LSTM) model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.

However, this model and most other models only consider a single emotion per speech sample and are often only trained on one or two databases. Kim et al. \cite{Kim2018a} is an exception as they consider the multi-label case of emotion recognition but they did not use machine learning and they used visual data on top of audio data. This paper extends upon the state-of-the-art by considering the problem of multi-label speech emotion classification using four databases to train and test the neural network.

@@

This paper is outlined as follows:
\begin{itemize}
	\item Section II details the methodology we use such as the how we combined the four databases and describes the architecture of the neural network.
	\item Section III describes the results from testing the neural network.
	\item Section IV discusses the results in the context of the field and how the results can be improved upon.
\end{itemize}

@@

We approach the problem of SER by preprocessing samples from four databases and then feeding those samples into a CNN. We preprocess the speech samples into log-Mel spectrogram to help the neural network extract features relevant to emotion recognition. This choice is based on previous work such as \cite{Engel2019} and \cite{Badshah2017}. The choice of using a CNN is based on previous work where Balakrishnan et al. \cite{Balakrishnan2017} found that CNN-based models have superior performance compared to RNNs. Another justification for using CNNs is due treating the log-Mel spectrograms as images where CNNs have been shown to perform well on.\cite{Krizhevsky2012}

A high level overview of the data flow is shown in figure \ref{highLevelDataFlowDiagram} where samples flow through the four stages of processing. Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} library respectively.

@@

For the STFT, we used a window size of 3072 with a 75\% overlap. This choice is based on the work of Zhao et al. \cite{Zhao2019} where they also use 75\% overlap but with a smaller window size (2048) and they achieved excellent results. For the Mel spectrogram, we set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins. The frequency range was chosen after experimenting with various ranges and selecting the range that resulted in visually clean spectrograms. The number of Mel bins was also chosen after experimentation as too few bins resulted in poor temporal resolution while too many bins resulted in poor frequency resolution. This is due to the Gabor limit.

\begin{figure}[]
	\centering
	\hspace{6mm}
	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
	\caption{A high level overview of the processing stages that a speech sample goes through.}
	\label{highLevelDataFlowDiagram}
\end{figure}

@@

After all of the databases were processed this way, the final combined database was fed into a neural network for training, validation, and testing. We constructed a eight layer neural network consisting of four convolutional layers and three dense layers. The network architecture is shown in figure \ref{neuralNetworkArchitecture}. The training process is described below.
\begin{itemize}
	\item We shuffled the combined database to make each input batch more uniform thus mitigating large gradient updates.
	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting. \cite{Srivastava2014}, \cite{Ioffe2015}
	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
	\item We used class weights during training to address class imbalance.
\end{itemize}

@@

The final model is evaluated on the testing set using two measures: categorical cross-entropy and accuracy. Categorical cross-entropy as defined in (\ref{cce}) where $N$ is the number of samples, $k$ is the number of classes, $t_{i,j}$ is 1 if the sample $i$ is in class $j$ and 0 otherwise, and $p_{i,j}$ is the model's predicted probability that sample $i$ is in class $j$.
\begin{equation}
\label{cce}
CCE = -\frac{1}{N}\sum_{i=1}^{N}\sum_{j=1}^{k}t_{i,j}log(p_{i,j})
\end{equation}
The choice to use categorical cross-entropy is based on its ease of use and how prevalent it is used in mutli-label problems.

The second measure, accuracy, is calculated from the confusion matrices shown in \ref{confusionMatrix}. The confusion matrices are calculated by comparing the true labels from the predicted labels for each emotion. There is a confusion matrix for each emotion due to the multi-label problem considering each emotion as independent. We define accuracy as the unweighted average accuracy of all emotions and this is mathematically described in (\ref{aveacc}) where $n$ is the total number of emotions (7 in this case), $TP_{i}$ is the number of true positives for emotion $i$, $TN_{i}$ is the number of true negatives for emotion $i$, and total is the total number of predictions.

@@

We chose this definition of accuracy so that we can compare our results to other papers that also use a similar measure.

@@

After training the model, it was evaluated on the testing set to get the final accuracy. The final accuracy achieved was 52.57\%. Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix}.

@@

We created a neural network model that achieved a 52.57\% accuracy on the problem of multi-label speech emotion classification. This is better than the state-of-the-art result of 52.14\% achieved in the single-label case. \cite{Zhao2019} So we conclude that it is possible to achieve a reasonable accuracy on the problem of multi-label speech emotion recognition by applying deep learning to the problem.

@@

Overall, this paper presented a 2D CNN model that achieved an accuracy of 52.57\% on the problem of multi-label speech emotion recognition using four combined databases. We obtained this result by transforming raw speech samples into log-Mel spectrograms using the STFT and the Mel scale. The log-Mel spectrograms are then fed into an eight layer neural network for classification. While this result is a promising start, we suggest improvements that future work can build upon to improve the accuracy of the model and to expand the scope of emotions considered.

>>>>

\begin{abstract}
Current speech emotion recognition systems often overlook the multi-label data that comes with databases. In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech. We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample. The model was able to classify the samples with an accuracy of 57.64\%. This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
\end{abstract}

@@

Current state-of-the-art SER has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN long short-term memory (LSTM) model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.

However, this model and most other models only consider a single emotion per speech sample and are often only trained on one or two databases. Kim et al. \cite{Kim2018a} is an exception as they consider the multi-label case of SER but they did not use machine learning and they used visual data on top of audio data. This paper extends upon current literature by considering the problem of multi-label SER using four databases to train and test a CNN.

@@

This paper is outlined as follows:
\begin{itemize}
	\item Section II details the methodology we use such as the how we combined the four databases and describes the architecture of the neural network.
	\item Section III describes and discusses the results from testing the neural network.
	\item Section IV concludes this paper and provides suggestions for future directions.
\end{itemize}

@@

We approach the problem of SER by preprocessing samples from four databases and then feeding those samples into a CNN. We preprocess the speech samples into log-Mel spectrogram to help the neural network extract features relevant to emotion recognition. This choice is based on previous work such as \cite{Engel2019} and \cite{Badshah2017}. The choice of using a CNN is also based on previous work where Balakrishnan et al. \cite{Balakrishnan2017} found that CNN-based models have superior performance compared to RNNs. Another justification for using CNNs is due treating the log-Mel spectrograms as images which CNNs have been shown to perform well on.\cite{Krizhevsky2012}

A high level overview of the data flow is shown in figure \ref{highLevelDataFlowDiagram} where samples flow through four stages of processing. The stages are detailed further into the methodology. Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} library respectively.

@@

For the STFT, we used a window size of 3072 with a 75\% overlap. This choice is based on the work of Zhao et al. \cite{Zhao2019} where they also use 75\% overlap but with a smaller window size (2048) and they achieved excellent results. For the Mel spectrogram, we set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins. The frequency range was chosen after experimenting with various ranges and selecting the range that resulted in visually clean spectrograms. The number of Mel bins was also chosen after experimentation as too few bins resulted in poor temporal resolution while too many bins resulted in poor frequency resolution. This is due to the Gabor limit which describes the trade-off between temporal and frequency resolution.

\begin{figure}[]
	\centering
	\hspace{6mm}
	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
	\caption{High level overview of the processing stages that a speech sample goes through.}
	\label{highLevelDataFlowDiagram}
\end{figure}

@@

After all of the databases were processed this way, the final combined database was fed into a neural network for training, validation, and testing. We constructed an eight layer neural network consisting of four convolutional layers and three dense layers. The network architecture is shown in figure \ref{neuralNetworkArchitecture}. The training process is described below.
\begin{itemize}
	\item We shuffled the combined database to make each input batch more uniform thus mitigating large gradient updates.
	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting. \cite{Srivastava2014}, \cite{Ioffe2015}
	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
	\item We used class weights during training to address class imbalance.
\end{itemize}

@@

The final model was evaluated on the testing set by using average accuracy. The average accuracy is calculated from the confusion matrices shown in figure \ref{confusionMatrix}. The confusion matrices are calculated by comparing the true labels to the predicted labels for each emotion. There is a confusion matrix for each emotion due to the multi-label problem considering each emotion as independent.

We define accuracy as the unweighted average accuracy of all emotions and this is mathematically described in (\ref{aveacc}) where $n$ is the total number of emotions (7 in our case), $TP_{i}$ is the number of true positives for emotion $i$, $TN_{i}$ is the number of true negatives for emotion $i$, and $Total_{i}$ is the total number of predictions for emotion $i$.

@@

In summary, we calculate the accuracy per emotion and then average over these accuracies to obtain the final accuracy. We chose this definition of accuracy so that we can compare our results to other papers that also use a similar measure of accuracy that is calculated from the confusion matrices.

@@

After training the model, it was evaluated on the testing set to get the final accuracy. The final accuracy achieved was 57.64\%. Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix} and comparisons to the current literature is shown in table \ref{litacccompare}.

@@

\bgroup
\def\arraystretch{2}
\begin{table}[]
	\centering
	\caption{Comparison of SER accuracy in literature.}
	\label{litacccompare}
	\begin{tabular}{|l|r|r|}
		\hline
		\multicolumn{1}{|c|}{\textbf{Research Work}} & \multicolumn{1}{c|}{\textbf{Method}} & \multicolumn{1}{c|}{\textbf{Accuracy (\%)}} \\ \hline
		\rowcolor[HTML]{F6F8FA}
		Zhao et al. \cite{Zhao2019} & CNN + LSTM & 52.1 \\ \hline
		Our Work & CNN & 57.6 \\ \hline
		\rowcolor[HTML]{F6F8FA}
		Etienne et al. \cite{Etienne2018} & CNN + LSTM & 61.7 \\ \hline
		Zhang et al. \cite{Zhang2019} & CNN & 63.9 \\ \hline
		\rowcolor[HTML]{F6F8FA}
		Fayek et al. \cite{Fayek2017} & CNN & 64.8 \\ \hline
		Yenigalla et al. \cite{Yenigalla2018} & CNN & 73.9 \\ \hline
		\rowcolor[HTML]{F6F8FA}
		Badshah et al. \cite {Badshah2019} & CNN & 84.3 \\ \hline
	\end{tabular}
\end{table}
\egroup

The results achieved in this paper do not reach the state-of-the-art accuracy achieved for the single-label case but this is the first attempt on the harder case of mutli-label using more data. So we conclude that it is possible to achieve a reasonable accuracy on the problem of multi-label speech emotion recognition by applying deep learning to the problem.

@@

Overall, this paper presented a 2D CNN model that achieved an accuracy of 57.64\% on the problem of multi-label speech emotion recognition using four combined databases. We obtained this result by transforming raw speech samples into log-Mel spectrograms using the STFT and the Mel scale. The log-Mel spectrograms are then fed into an eight layer neural network for classification. While this result is a promising start, we suggest improvements that future work can build upon to improve the accuracy of the model and to expand the scope of emotions considered.
